---
title: "Clustering and PCA"
subtitle: "Data Science for Biologists, Spring 2020"
author: "Poonam Shah"
output: 
  html_document:
    highlight: tango
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

## SET YOUR SEED BELOW!!! Make it YOUR SEED! Uncomment the line, and add your chosen seed number into the parentheses
set.seed(70)

```

## Instructions

Standard grading criteria apply. And...

+ Make sure you set your seed
+ Proofread to submit YOUR OWN WORDS!!
+ Make sure all bullets properly render in the knitted HTML. **YOU WILL LOSE POINTS FOR BULLETED LISTS THAT DO NOT TURN OUT PROPERLY IN THE HTML!!** 
+ Remember that `#` do NOT make comments in markdown. They make HEADERS!! You need to use HTML comments to have comments in your text that don't appear in the knitted HTML. For example, the line below does not appear in html.

<!-- This is an actual comment in markdown. -->

## The data

We will be using a dataset of olives collected from various regions of Italy. The data gives the percentage of different fatty acids that comprise the oil in the olives. [Source](https://rdrr.io/cran/dslabs/man/olive.html)

```{r, collapse=T}
olives <- read_csv("https://raw.githubusercontent.com/sjspielman/datascience_for_biologists/master/data/olives.csv")
dplyr::glimpse(olives)
```


## Part 1: Clustering

Perform TWO k-means cluster analysis, once with **k=3** and once with **k=5**. 

```{r}
## Perform two clustering analyses here with k=3 and k=5, separately.


#k=3 clustering
olives %>%
  select(-region, -area) %>%
  kmeans(3) -> olives_k3


#k=5 clustering
olives %>%
  select(-region, -area) %>%
  kmeans(5) -> olives_k5
```



For each analysis, make a SINGLE FIGURE for each clustering analysis (two figures total). Choose your own adventure for each figure! You should explain per figure: a) Why did you decide to visualize the clustering using this particular figure? (~1 sentence) and b) What insights (if any!) does this figure provide (1-2 sentences)? Answer in the bullets below the given figure. 

**You should not make two identical figures - they need to be _somewhat_ different.** For example, two barplots showing the relationship between clusters and regions is a no-go. Two scatterplots but with different x-axis and y-axis variables on the axes is ok! Two boxplots (violin, den sity, histogram, sina, jitter...) but with a different focal variable is ok! Either way, your figure MUST show the clusters.

```{r}
## Figure for k=3 including all associated wrangling
olives %>%
  mutate(k3_clusters = factor(olives_k3$cluster)) %>%
  ggplot(aes(x=palmitic, fill=region))+
  geom_density(position = "stack", alpha=0.4) +
  #picked the color
  scale_fill_brewer(palette = "Dark2") +
  #clean up the axis
  labs(x="Palmitic", y="Density") +
  theme_bw()


```

+ **Answer: Why did you choose this figure?**
+ I chose the density plot it will show the best to see the Palmitic in the different regions

+ **Answer: What trends/insights (if any)  does the figure reveal, as they relate to the clusters?**
+ The pot hsows the Palmitic is more in density in the Northern Italy region. The Southern region grows Palmitic higher in numbers but not in density. 


```{r}
## Figure for k=5 including all associated wrangling
olives %>%
  mutate(k3_clusters = factor(olives_k3$cluster)) %>%
  #Oleic vs Linoeic
  ggplot(aes(x=oleic, y=linoleic, color=area)) +
  geom_point() +
  facet_wrap(~area, scales = "free_y", nrow = 3) +
  #changing the colors
  scale_color_brewer(palette = "Paired") +
  #cleaning up the axis
  labs(x="Oleic", y="Linoleic") +
  theme_light()

```

+ **Answer: Why did you choose this figure?**
+ The point plot is showing the clusters of the two different type of olives species based on area.  
+ **Answer: What trends/insights (if any) does the figure reveal, as they relate to the clusters?**
+ The graphs look like its going to have a negative slope. The Inland-Sardinia has a slope of 90 angles.


## Part Two: PCA

Perform a principle components analysis on this dataset, and create TWO figures from your PCA. Each figure should have a brief explanation as a bullet below it. For each figure, your should be labeled to include information about how much variation the given PC explains in the data.

```{r}
## Code to perform PCA goes here
olives %>%
  select(-region, -area) %>%
  scale() %>%
  prcomp() -> olives_pca

```

1. But first, use code to answer this question! Intepret how variables in the initial dataset **load** on the PCs by examining, you guessed it, the loadings. In the bullets below, state which SINGLE variable load most strongly on PC1 and PC2 each, and which SINGLE variable loads most weakly on PC1 and PC2.

```{r}
## Code to examine loadings goes here
olives_pca$rotation
```

+ **Strongest PC1 loading: oleic**
+ **Strongest PC2 loading: linolenic**
+ **Weakest PC1 loading: stearic**
+ **Weakest PC2 loading: palmitics**

2. Plot PC2 against PC1 (PC2 is on Y and PC1 is on X) as a scatterplot, where you use color to distinguish among **olive region** (NOT area!). Make sure to show ellipses for each grouping. Do NOT include loading arrows. In the bullet below, _explain in 2-3 sentences_ if and how either PC1 or PC2 discriminates olive region. You may write a bulleted list *as long as it renders properly.*

```{r}
## Code to make this figure goes here
as_tibble(olives_pca$x) %>%
  mutate(region = olives$region) %>%
  ggplot(aes(x=PC1, y= PC2, color=region)) +
  geom_point() +
  #circles the region
  stat_ellipse() +
  #changing the color of the points
  scale_color_brewer(palette = "Spectral") +
  #changing background to a grey
  theme_dark()


```

+ **Explanation for figure**
+ It does look like the PC1 is discriminated in the olive region. 
+ Most of the points separated and there is not a lot of overlapping. 

2. Plot PC2 against PC1 as a scatterplot, where you use color to distinguish among **olive area** (the other categorical column!). Make sure to show ellipses for each grouping. Do NOT include loading arrows. In the bullet below, _explain in 2-3 sentences_ if and how either PC1 or PC2 discriminates olive region. You may write a bulleted list *as long as it renders properly.*
```{r}
as_tibble(olives_pca$x) %>%
  mutate(area = olives$area) %>%
  ggplot(aes(x=PC1, y= PC2, color=area)) +
  geom_point() +
  #circles the area
  stat_ellipse()+
  #changing the color of the points
  scale_color_brewer(palette = "Paired") +
  theme_bw()

```

+ **Explanation for figure**
+ It does not look like PC1 nor PC2 is being discriming over an area.
+ All of the circles are over lapping with each area. 


