---
title: "ANSWERS: Working with untidy and tidy data"
subtitle: "Data Science for Biologists, Spring 2020"
author: "Poonam Shah"
output: 
  html_document:
    highlight: espresso
css: static/sjs-style.css
params:
  data_path: "data/"
editor_options: 
  chunk_output_type: console
---

<!-- SETUP CODE CHUNK -->
```{r setup, include=FALSE}
## Change the default figure width and height to your liking, but within reason.
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3.5)
library(tidyverse)
library(kableExtra) # helpful to print fancy tables in RMarkdown
theme_set(theme_minimal())  ### If you want to set a global theme, do so here


color_choice_1="coral"
color_choice_2="darkslategray"
```


# <mark>Questions</mark>

#### Preface Question

Look at the YAML front matter: there is something NEW! Read about params [here](https://rmarkdown.rstudio.com/lesson-6.html). In this case we set a global parameter for the RMarkdown file that corresponds to the path where data lives! We can proceed to use this as a variable like so (always `params$`, like a column in a data frame!). Use this variable for your paths, e.g. `read_csv(paste0(params$data_path, <NAMEOFFILE>))`
```{r Preface Question}

params$data_path


```

#### Question 1

```{r Question 1 part 1}
# Code chunk to read in and examine the data with head()

measle <- read_csv(paste0(params$data_path, "mcv_immunized_percent_of_one_year_olds.csv"))

head(measle)
```


```{r Question 1 part 2}
# Code chunk to convert data to tidy
#data and piped into the function
measle %>% 
  #pivot_longer untidy to tidy. minus country
  pivot_longer(-country, names_to ="Year", values_to="Percent") -> Measle_data_from_q1 #new variable

head(Measle_data_from_q1)

```
#### Question 2


```{r Question 2}
# Code chunk to plot the data
#dataset piped
Measle_data_from_q1 %>%
  #filter and using an array to show both country
  filter(country %in% c("Argentina", "Dominica")) %>%
  #setting up the axis and groups for the line plot
  ggplot(aes(x=Year, y=Percent, color=country, group=country)) +
  #line graph
  geom_line()+
  #the points graph
  geom_point()+
  #clear up the axis
  labs(x="Year", y="Percent")+
  #changed the colors
  scale_color_manual(values = c(color_choice_1,color_choice_2))+
  #changed the angle of the year
  theme(axis.text = element_text(angle = 90))


```

<answer>
In a few sentences below the figure, discuss the trends you observe.

Dominica started the measle vaccines in 1982 it started even out around 1993. Argentina started measle at 1980. Argentina is uneven in the measle vaccines. 

</answer>

#### Question 3


```{r Question 3 part 1}
# Code chunk to read in and examine the data with head()
#reading in the code
Diphtheria_tetanus_pertussis <- read_csv(paste0(params$data_path, "dtp3_immunized_percent_of_one_year_olds.csv"))

#Showing the data
head(Diphtheria_tetanus_pertussis)

```


```{r Question 3 part 2}
# Code chunk to convert data to tidy and combine with question 1 data

#tidying up Diphtheria-tetanus-pertussis
Diphtheria_tetanus_pertussis %>%
pivot_longer(-country, names_to ="Year", values_to="Percent") %>%
  #adding a new column
  mutate(Vaccine = "Diphtheria-tetanus-pertussis" ) -> DTP_tidy

#Adding a new column
Measle_data_from_q1 %>%
  mutate(Vaccine = "Measles") -> Measle_tidy_q3

#combing the dataset
bind_rows(DTP_tidy,Measle_tidy_q3) -> DTP_Mealse_data

#changing year to a char to a dbl
DTP_Mealse_data %>%
mutate(Year = as.numeric(Year)) %>%
#filtering between 2005-2011
filter(between(Year, 2005,2011)) -> DTP_Mealse_2005_2011_data

```

```{r Question 3 part 3}
# Code chunk to plot the data
DTP_Mealse_2005_2011_data %>%
  #filter and using an array to show both country
  filter(country %in% c("India", "France")) %>%
  #setting up the axis and groups for the line plot
  ggplot(aes(x=Year, y=Percent, color=Vaccine, group=country)) +
  #line graph
  geom_line()+
  #the points graph
  geom_point() +
  xlim(2005, 2011) +
  facet_wrap(Vaccine~country) +
  scale_color_manual(values = c(color_choice_1,color_choice_2)) +
  theme(legend.position="none")

```

<answer>
France had a better vaccine rate then India. France is a first world country it would have a better medicine then India.
India has a low vaccine rate for both Measles and DTP.  
</answer>

#### Question 4


```{r Question 4 part 1}
# Code chunk to read in and examine the data with head()

#reading in the data
seals<-read_csv(paste0(params$data_path, "seals.csv"))
#viewing the data
head(seals)

```


```{r Question 4 part 2}
# Density plot with feeding
seals %>%
  #x axis and fill with feeding
  ggplot(aes(x=oxygen_use, fill=feeding)) +
  #density plot
  geom_density(alpha=0.6) +
  #cleaning up the axis
  labs(x="Oxygen usage", y="Density") +
  scale_fill_manual(values = c(color_choice_1,color_choice_2))

```

```{r Question 4 part 3}
# Code chunk for data un-tidying and scatterplot
seals %>%
  pivot_wider(names_from = feeding, values_from = oxygen_use) %>%
  #renaming the columns
  rename(oxygen_use_not_feeding = No) %>%
  rename(oxygen_use_feeding_on = Yes) -> seals_untidy
  
  
seals_untidy %>%  
  ggplot(aes(x=oxygen_use_not_feeding, y=oxygen_use_feeding_on)) +
  geom_point(aes( color=seal)) +
  # add reference lines
  geom_abline() + 
  #fixing the axis units
  xlim(30,120) +
  ylim(0,150) +
  #making sure every point has a different color
  scale_color_distiller(palette = "Dark2") +
  #clearing up the axis
  labs(x="oxygen use not feeding", y="oxygen use feeding on")

```

<answer>
In a few sentences below the figure, discuss whether EACH FIGURE shows evidence for the researchers’ prediction.

The first graph is a simple density plot. It is hard to read the graph. It looks like the seals with no feeding had a highest oxygen usege. The second graph is a scatterplot. It shows the Qxygen use feeding on has a better oxygen level. It shows a postive trendline
</answer>

#### Question 5


```{r Question 5 part 1}

#reading in the data
planets <- read_csv2(paste0(params$data_path, "planets.csv"))
#examine the data with head()
head(planets)
```


```{r Question 5 part 2}

planets %>% 
  #separate the items in the terrain column
  separate_rows(terrain) %>%
  #separate the items in the climate column
  separate_rows(climate) -> planets_tidy

```


#### Question 6

```{r Question 6}

planets_tidy %>%
  #separated the date and time from T
  separate(edited, c("edited_date", "edit_time"), sep="T") %>%
  #uniting the two columns together
  unite(ratio_period_rotation_orbital, rotation_period, orbital_period) %>%
  #uniting the two columns together
  unite(name_pop, name,population) %>%
  #only contains mountains from terrain
  filter(terrain=="mountains") %>%
  arrange(desc(ratio_period_rotation_orbital))-> planets_q6

```


#### Question 7


```{r Question 7 part 1}
## defines each column as an array
tibble(
  color        = c("blue", "green", "turquoise"),
  male         = c(25, 36, 52),
  female       = c(28, 17, 33),
  species_abbr = c("sp1", "sp1", "sp1")
) -> frogs_sp1

## defines tibble in a format like the tibble would end up appearing
## spacing is helpful!
tribble(
  ~color, ~male, ~female, ~species_abbr,
  "orange", 13,   17,    "sp2",
  "red",    45,   36,    "sp2",
  "pink",  31,    25,    "sp2"
) -> frogs_sp2

## Like strategy one
tibble(
  species_name = c("Frogus firstus", "Frofus secondus"),
  species_abbr = c("sp1", "sp2")
) -> species_names

### ADD CODE HERE TO PRINT OUT THESE THREE DATASETS!!! YOU HAVE TO LOOK AT THEM TO USE THEM!!
print(frogs_sp1)
print(frogs_sp2)
print(species_names)

```



```{r Question 7 part 2}
# Code chunk to tidy and combine the three datasets
bind_rows(frogs_sp1,frogs_sp2) ->full_frogs

left_join(full_frogs,species_names) ->full_frogs_species

full_frogs_species %>%
  #separate species name to genus and species
  separate(species_name, c("genus", "species")) %>%
  #deleted the species_abbr
  select(-species_abbr) %>%
  #copying the male column
  mutate(number=male) %>%
  #copying the female column
  mutate(NUMBER=female) %>%
  #combine the number columns
  unite(Numbers, number, NUMBER) %>%
  #separating the data in the column
  separate_rows(Numbers) %>%
  #changing male from values to male
  mutate(Male_sex= if_else(male >=5, "male", "female")) %>%
  #changing female from values to female
  mutate(Female_sex= if_else(female>=5, "female", "male")) %>%
  #combine the genders
  unite(Sex, Male_sex, Female_sex) %>%
  #separate in the row
  separate_rows(Sex) %>%
  #delecting the old male and female
  select(-male,-female) %>%
  #keeping the correct rows
  slice(1,4,5,8,9,12,13,16,17,20,21,24)->final_frog

```

```{r Question 7 part 3}

final_frog %>%
  #filtering it by color
  filter(color %in% c("blue", "green")) %>%
  #show the female
  filter(Sex=="female")

#There are more blue frugus firstus 
   
```

```{r Question 7 part 4}

final_frog %>%
#filtering it by male
filter(Sex=="male") %>%
#show the most to the least by numbers column
arrange(desc(Numbers))

#The Forgies firstus has the most male frogs
```

```{r Question 7 part 5}
final_frog %>%
  #making the tidy to untidy
  pivot_wider(names_from = Sex, values_from = Numbers) %>%
  #filter the species
  filter(species=="secondus") %>%
  #changing the columns to chr to a numeric
  mutate(male = as.numeric(male)) %>%
  mutate(female=as.numeric(female)) %>%
  #the ratio formula
  mutate(ratio=male/female)
 
```



#### Question 8

```{r Question 8}
# Code chunk for kable()

final_frog %>%
  kable()

final_frog %>%
  kable() %>%
  kable_styling("striped", full_width = F) %>%
  row_spec(1:2, bold = T, color = "blue", background = "white") %>%
  row_spec(3:4, bold = T, color = "green", background = "white") %>%
  row_spec(5:6, bold = T, color = "darkturquoise", background = "white") %>%
  row_spec(7:8, bold = T, color = "orange", background = "white") %>%
  row_spec(9:10, bold = T, color = "red", background = "white") %>%
  row_spec(11:12, bold = T, color = "pink", background = "white")
  

final_frog %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", font_size = 7)


```

#### BONUS QUESTION

*Add your own chunks if you choose to do this adventure.*
```{r Bonus Question}

```
<answer>
Replace this text with your answer.
</answer>